---
title: "Lab09"
author: "Marc Brooks"
date: "11/2/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```


```{r packages}
library(tidyverse)
library(rstanarm)
library(magrittr)
library(ggplot2)
library(mlmRev)
library(tidybayes)
library(ggstance)
library(dplyr)
library(modelr)
```




```{r}
data(Gcsemv, package = "mlmRev")
dim(Gcsemv)

summary(Gcsemv)


# Make Male the reference category and rename variable
Gcsemv$female <- relevel(Gcsemv$gender, "M")

# Use only total score on coursework paper
GCSE <- subset(x = Gcsemv,
               select = c(school, student, female, course))

# Count unique schools and students
m <- length(unique(GCSE$school))
N <- nrow(GCSE)
```

# Exercise 1

```{r ex1, fig.height=4, fig.width=6}
GCSE %>% 
  group_by(school) %>% 
  summarise(mean_score = mean(course)) %>% 
  ggplot(aes(x=mean_score)) +
  geom_histogram(binwidth = 2, fill = "#3182bd", alpha=.7) +
  theme_classic() +
  labs(x="Sample Means")
```


```{r}
pooled <- stan_glm(course ~ 1 + female, data = GCSE, refresh = 0)
unpooled <- stan_glm(course ~ -1 + school + female,data=GCSE, refresh = 0)
```

```{r}
mod1 <- stan_lmer(formula = course ~ 1 + (1 | school),
                  data = GCSE,
                  seed = 349,
                  refresh = 0)
```

```{r}
summary(mod1,
        pars = c("(Intercept)", "sigma", "Sigma[school:(Intercept),(Intercept)]"),
        probs = c(0.025, 0.975),
        digits = 3)
```

# Exercise 2

The posterior mean for $\mu_{\theta}$ is 73.704 and the 95% Credible Interval is $[71.531, 75.935]$.
The posterior mean for $\sigma$ is 13.816 and the 95% Credible Interval is $[13.351, 14.308]$.
The posterior mean for $\tau^2$ is 79.674 and he 95% Credible Interval is $[54.390, 115.156]$.


```{r}
mod1_sims <- as.matrix(mod1)
dim(mod1_sims)
```

```{r}
par_names <- colnames(mod1_sims)
head(par_names)
```


```{r}
tail(par_names)
```

```{r}
# obtain draws for mu_theta
mu_theta_sims <- as.matrix(mod1, pars = "(Intercept)")
```

```{r}
# obtain draws for each school's contribution to intercept
theta_sims <- as.matrix(mod1,
                        regex_pars ="b\\[\\(Intercept\\) school\\:")
```


```{r}
# to finish: obtain draws for sigma and tau^2
sig_sims <- as.matrix(mod1,
                      pars = "sigma")
tau2_sims <- as.matrix(mod1,
                       pars = "Sigma[school:(Intercept),(Intercept)]")
```

```{r}
int_sims <- as.numeric(mu_theta_sims) + theta_sims

# posterior mean
int_mean <- apply(int_sims, MARGIN = 2, FUN = mean)

# credible interval
int_ci <- apply(int_sims, MARGIN = 2, FUN = quantile, probs = c(0.025, 0.975))
int_ci <- data.frame(t(int_ci))

# combine into a single df
int_df <- data.frame(int_mean, int_ci)
names(int_df) <- c("post_mean","Q2.5", "Q97.5")

# sort DF according to posterior mean
int_df <- int_df[order(int_df$post_mean),]

# create variable "index" to represent order
int_df <- int_df %>% mutate(index = row_number())

# plot posterior means of school-varying intercepts, along with 95 CIs
ggplot(data = int_df, aes(x = index, y = post_mean))+
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5))+
  scale_x_continuous("Index", breaks = seq(0,m, 5)) +
  scale_y_continuous(expression(paste("varying intercept ", theta[j])))
```

Let's look at two schools with small sample sizes.
```{r}
GCSE %>% 
  group_by(school) %>% 
  summarise(count  = n()) %>% 
  arrange(count) %>% 
  slice(1:10)
  
```

```{r}
theta_simsComp <- as.matrix(mod1,
                        regex_pars ="b\\[\\(Intercept\\) school\\:(25241|22908)")
```

```{r}
print(paste("Posterior mean difference of average scores =  = ", round(mean(theta_simsComp[,1] - theta_simsComp[,2]),4)))

CI <- quantile(round(theta_simsComp[,1] - theta_simsComp[,2],4), c(0.025, .975))
print(paste("95% CI = ", '[',  str_c(CI, sep=","), ']'))

data.frame(diff = theta_simsComp[,1] - theta_simsComp[,2]) %>% 
  ggplot(aes(x=diff)) +
  geom_histogram(binwidth = 2, fill = "#3182bd", alpha=.7) +
  theme_classic() +
  labs(x="Difference in posterior means for 25242 and 22908")
```

```{r}
mod2 <- stan_lmer(formula = course ~ 1 + female + (1 | school),
                  data = GCSE, 
                  prior = normal(location = 0,
                                        scale = 100,
                                        autoscale = F),
                  prior_intercept = normal(location = 0,
                                        scale = 100,
                                        autoscale = F),
                  seed = 349,
                  refresh = 0)
```

```{r}
# plot varying intercepts
mod2.sims <- as.matrix(mod2)
group_int <- mean(mod2.sims[,1])
mp <- mean(mod2.sims[,2])
bp <- apply(mod2.sims[, 3:75], 2, mean)
xvals <- seq(0,1,.01)
plot(x = xvals, y = rep(0, length(xvals)), 
     ylim = c(50, 90), xlim = c(-0.1,1.1), xaxt = "n", xlab = "female", ylab = "course")
axis(side = 1, at = c(0,1))
for (bi in bp){
  lines(xvals, (group_int + bi)+xvals*mp)
}
```

# Exercise 4
```{r}
summary(mod2,
        pars = c("(Intercept)", "femaleF", "sigma", "Sigma[school:(Intercept),(Intercept)]"),
        probs = c(0.025, 0.975),
        digits = 3)
```

The posterior mean for $\mu_{\theta}$ is 69.669 and the 95% Credible Interval is $[67.322, 72.056]$.
The posterior mean for $\beta$ is 6.744 and the 95% Credible Interval is $[5.410, 8.043]$.
The posterior mean for $\sigma$ is 13.424  and the 95% Credible Interval is $[12.965, 13.905]$.
The posterior mean for $\tau^2$ is 80.663 and he 95% Credible Interval is $[54.072, 118.776]$.


# Model 3


```{r}
mod3 <- stan_lmer(formula = course~ 1+ female + (1 + female | school),
                  data = GCSE,
                  seed = 349,
                  refresh = 0)

mod3_sims <- as.matrix(mod3)
```


```{r}
# obtain draws for mu_theta
mu_theta_sims <- as.matrix(mod3, pars = "(Intercept)")

fem_sims <- as.matrix(mod3, pars = "femaleF")
# obtain draws for each school's contribution to intercept
theta_sims <- as.matrix(mod3,
                        regex_pars ="b\\[\\(Intercept\\) school\\:")
beta_sims <- as.matrix(mod3,
                       regex_pars ="b\\[femaleF school\\:")

int_sims <- as.numeric(mu_theta_sims) + theta_sims
slope_sims <- as.numeric(fem_sims) + beta_sims

# posterior mean
slope_mean <- apply(slope_sims, MARGIN = 2, FUN = mean)

# credible interval
slope_ci <- apply(slope_sims, MARGIN = 2, FUN = quantile, probs = c(0.025, 0.975))
slope_ci <- data.frame(t(slope_ci))

# combine into a single df
slope_df <- data.frame(slope_mean, slope_ci, levels(GCSE$school))
names(slope_df) <- c("post_mean","Q2.5", "Q97.5", "school")

# sort DF according to posterior mean
slope_df <- slope_df[order(slope_df$post_mean),]

# create variable "index" to represent order
slope_df <- slope_df %>% mutate(index = row_number())

# plot posterior means of school-varying slopes, along with 95% CIs
ggplot(data = slope_df, aes(x = index, y = post_mean))+
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5))+
  scale_x_continuous("Index", breaks = seq(1,m, 1),
                     labels = slope_df$school) +
  scale_y_continuous(expression(paste("varying slopes ", beta[j])))+
  theme(axis.text.x = element_text(angle = 90))
```


# Model Comparison

```{r}
loo1 <- loo(mod1)
loo2 <- loo(mod2)
loo3 <- loo(mod3)

loo_compare(loo1,loo2,loo3)
```

```{r}
pooled.sim <- as.matrix(pooled)
unpooled.sim <- as.matrix(unpooled)
m1.sim <- as.matrix(mod1)
m2.sim <- as.matrix(mod2)
m3.sim <- as.matrix(mod3)
schools <- unique(GCSE$school)


alpha2 = mean(m2.sim[,1])
alpha3 <- mean(m3.sim[,1])

partial.fem2 <- mean(m2.sim[,2])
partial.fem3 <- mean(m3.sim[,2])
unpooled.fem <- mean(unpooled.sim[,74])

par(mfrow = c(2, 3), mar = c(1,2,2,1))
for (i in 1:18){
  temp = GCSE %>% filter(school == schools[i]) %>%
    na.omit()
  y <- temp$course
  x <- as.numeric(temp$female)-1
  plot(x + rnorm(length(x)) *0.001, y, ylim = c(35,101), xlab = "female",main =schools[i], xaxt = "n", ylab = "course")
  axis(1,c(0,1),cex.axis=0.8)
  
  # no pooling
  b = mean(unpooled.sim[,i])

  # plot lines and data
  xvals = seq(-0.1, 1.1, 0.01)
  lines(xvals, xvals * mean(pooled.sim[,2]) + mean(pooled.sim[,1]), col = "red") # pooled
  lines(xvals, xvals * unpooled.fem + b, col = "blue") # unpooled
  lines(xvals, xvals*partial.fem2 + (alpha2 + mean(m2.sim[,i+2])) , col = "green") # varying int
  lines(xvals, xvals*(partial.fem3 + mean(m3.sim[, 2 + i*2])) + (alpha3 + mean(m3.sim[, 1 + i*2])), col = "orange") # varying int and slope
  legend("bottom", legend = paste("n =", length(y), " "))
}
```

# Exercise 5